Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-12-09 paper.tex
text all updated

Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
The size and location of tunnels determine the ligand size and shape which can enter the active site.
Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
This helps them to reduce the number of necessary in-vitro lab experiments.
The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
\textcolor{blue}{However, the ligands are typically of a non-spherical shape which is completely ignored by these methods. 
Our approach aims to overcome this problem by taking into account the actual shape of the ligand and additionally also its evolution over time, i.e. its flexibility.}

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{-0.5}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{ccc}
\includegraphics[width=0.14\textwidth]{fig/motiv1} &
\includegraphics[width=0.15\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.14\textwidth]{fig/motiv3}  \\
Protein 1CQW & Active site & Detected tunnels \\ %& Example of a  \\
             &            & (orange)         \\  %& trajectory
\multicolumn{3}{c}{%
\includegraphics[width=0.16\textwidth]{fig/renderDCP}  \hskip 15pt
\includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
\end{tabular}
}
\caption{\label{fig::motiv}
    \small
    Tunnels in the DhaA haloalkane dehalogenase protein (PDB ID 1CQW) with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
}
\end{figure}

\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.
We can formulate this task as a motion planning problem and employ sampling-based methods to solve it.
%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.

Sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples
into a graph structure (roadmap).
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
pathway detection~\cite{cortes2010simulating,cortes2005path} and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving

@@

Finding trajectories only along a given tunnel tunnel is faster than computing any exit-pathway from the protein due to decreased
volume of the configuration space to be searched.
\textcolor{blue}{Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the outside solvent towards the protein active site.}

Our proposed method was tested by the biochemists who compared the results with already known trajectories of ligands, determined by their traditional approaches.
\textcolor{blue}{In the following we first mention several approaches and methods related to our solution and then describe our proposed method in detail.}

@@

{\textcolor{blue}{The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using
    sampling-based motion planners.
These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider arbitrary geometry.}
Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
%Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
%The computation of the ligand trajectories can be solved also using motion 
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g.
by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identifed using watershed labeling.
The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
%The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
%based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using Probabilistic Roadmaps.
Each configuration in the roadmap is also evaluated using electorstatic energy.
The high-energy configurations are discared, while low-energy conformations (and their vicinity) are sampled more dense.
The approach~\cite{singhLIG} can be also used to predict ligand binding sites.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
A random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it can be considered as a new starting configuration and it is added to $\QI$.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

@@

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, 
and the rotational part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The ligand index $l$ and scale $s$ can be set to zero, as these are not used in the employed metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
We propose to set this parameter to the average width of the tunnel.
%By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
%of $\rv$ cause the exploration of paths around the tunnel.

@@

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
This metric is sensitive to the shape of the ligand (that is given by it's conformation) as it computes the distance using the two nearest atoms of the two configurations.

By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
is useful when multiple conformations are utilized.
This metric ensures that the tree is expanded by such configurations, whose atoms maximally approach
the atoms of the configuration $\qrand$.
We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability
of reaching the active site.

@@

The pool of available conformations $\L$ consists all conformations (upto some resolution) with energy lower than a user speficied threshold, which can still contain too many conformations (thousands or even more).
To select a subset of conformation, the distance of the ligand atoms to the geometric center of the ligand is computed and 
deviation of these distances is calculated.
Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.

@@

The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per configuration, while \RA\ extends the tree only by one configuration.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 0.5~\AA.

In the following experiments, we measured the performance of the methods using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

The performance of \RA\ and \RB\ planners is first measured by finding trajecotries for only one atom of VDW radii $1.7$~\AA.
The \RC\ variant is not tested here as it behaves same as \RA\ in the case of one-atom ligand.
The computed traversability rates are shown in Tab.~\ref{tab::one}.
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
The row 'Planning' shows the traversable rate computed by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increase the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphSel.tex
}
\end{table}

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    Each frame of protein dynamics consists of 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphNew.tex
}
\end{table}

@@

The traversability rate is shown in Tab.~\ref{tab::main}. 
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the ligand DCP, trajectories were detected in 10~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promissing.
However, their binding was not approved yet by MD dedicated simulations.

@@

Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
to distance $\dt=2$~\AA and they were considered unsuccessful otherwise.
The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
also in other places.
The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
The detail is depicted in Fig.~\ref{fig::detail}.
Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.

>>>>

Reactivity of proteins with other molecules plays a crucial role in many research disciplines, including protein engineering and drug design.
Protein reactivity is tightly connected with the presence of the void space between its amino acids. 
This space, forming a path called tunnel, can be utilized by a small molecule (ligand) which can enter the protein inner space and react with its amino acids surrounding a specific site, called the active site~\cite{gora2013gates}.
The size and location of tunnels determine the ligand size and shape which can enter the active site.
Currently the biochemists are using various computational methods for tunnel detection (Fig.~\ref{fig::motiv}) in order to estimate the possibility of interactions between a protein and a ligand.
This helps them to reduce the number of necessary in-vitro lab experiments.
The most common methods detect the tunnels using Voronoi diagrams~\cite{yaffe2008,caver3}, assuming a spherical ligand (probe), i.e., the approximation of the ligand by a bounding sphere.
However, the ligands are typically of a non-spherical shape which is completely ignored by these methods.

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{-0.5}
\renewcommand{\tabcolsep}{-1pt}
\begin{tabular}{ccc}
\includegraphics[width=0.14\textwidth]{fig/motiv1} &
\includegraphics[width=0.15\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.14\textwidth]{fig/motiv3}  \\
Protein  & Active site & Detected tunnels \\ %& Example of a  \\
             &            & (orange)         \\  %& trajectory
\multicolumn{3}{c}{%
\includegraphics[width=0.16\textwidth]{fig/renderDCP}  \hskip 15pt
\includegraphics[width=0.14\textwidth]{fig/render37t}} \\ 
\end{tabular}
}
\caption{\label{fig::motiv}
    \small
    Tunnels in the Haloalkane dehalogenase protein with possible trajectories of 1-Chlorpropan ligand (bottom left) and 1,2-Dichloroethane (bottom right).
}
\end{figure}

Our approach aims to overcome this problem by taking into account the actual shape of the ligand.
This necessary requires to consider also conformational changes.
%Our approach aims to overcome this problem by taking into account the actual shape of the ligand and also its conformations.
%\textcolor{blue}{To provide the biochemists with a solution which detects tunnels also in these frames and understand the ligand passage, it is necessary to take into account not only the ligand shape but also its conformation changes when the ligand moves.}
%The proposed solution needs also to cope with the narrow passage problem, as proteins are dense structures limiting the ligand movements inside.
Such a ligand, changing the mutual position of its atoms within a molecular dynamics, is denoted as a flexible ligand.

The analysis of traversability of the tunnels by flexible ligands can be formulated as a motion planning problem and
solve by sampling-based planners.
Sampling-based planners randomly sample the configuration space of the robot (ligand) and connect the collision-free samples
into a graph structure (roadmap).
Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT} is one of the most used sampling-based planners.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
e.g., to study loop motions~\cite{cortes2004geometric}, to identify binding sites~\cite{bayazit2001ligand}, for exit
pathway detection~\cite{cortes2010simulating,cortes2005path} and especially for protein folding~\cite{raveh2009rapid,amato2002using}. %novinskaya2015improving

@@

Finding trajectories only along a given tunnel tunnel is faster than computing any exit-pathway from the protein due to decreased
volume of the configuration space to be searched.
Another advantage of our method in comparison with the existing approaches is that we enable the users to trace the ligand path leading from the outside solvent towards the protein active site.

Our proposed method was tested by the biochemists who compared the results with already known trajectories of ligands, determined by their traditional approaches.
In the following we first mention several approaches and methods related to our solution and then describe our proposed method in detail.

@@

The traversability of a tunnel by a flexible ligand can be formulated as a motion planning problem and solved using sampling-based motion planners.
These planners can cope with objects (robots) with many Degrees of Freedom (DOF)  and they can also consider arbitrary geometry.
Two widely used planners are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Tree (RRT)~\cite{lavalleRRT}.
%Sampling-based planners like RRT can be used to find the trajectories even for flexible ligands, as they can cope with many-DOF systems and they can consider arbitrary shape.
%The computation of the ligand trajectories can be solved also using motion 
%In order to determine if the ligand can pass the tunnel using the geometric-based methods, it is necessary to compute a trajectory considering the shape of the ligand and its 3D translation and rotation.
%In each iteration of RRT, a random configuration $\qrand \in \C$ is generated and its nearest node $\qnear \in \T$ in the tree is found.
%A new configuration $\qnew$ is constructed on the line connecting $\qnear$ and $\qrand$ in the distance $\varepsilon$ from $\qnear$.
%If $\qnew$ is collision-free, it is added to the tree.
%The algorithm terminates if the tree approaches the goal configuration close enough.
%rrt for molecular: \cite{al2012motion}
%survey \cite{gipson2012computational}
%Solutions for tunnel detection using sampling-based methods however have not been discussed yet.
%Sampling-based planners have been used in various applications in robotics~\cite{elbanhawi2014sampling} and also  %,latombe1999motion} and also
% to study proteins, e.g. in  
%loop motions~\cite{cortes2004geometric},
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
%protein folding~\cite{raveh2009rapid,novinskaya2015improving},
%or protein folding combined with ligand diffusion~\cite{cortes2010simulating}.
The well known issue of the sampling-based planners is the narrow passage problem.
Narrow passage is a small region in the configuration space, whose removal changes the connectivity of the space.
Due to its low volume, the probability of sampling the narrow passage is low.
Consequently, many iterations are needed in order to put enough samples into the passages, which increases the computation time.
PRM-based planners can cope with the narrow passage problem by increasing the probability of sampling in difficult regions, e.g.
by sampling along medial axis~\cite{amatoOBRRT,amato2002using,wilmarthMAPRM}.
In~\cite{bergWIG}, workspace is represented by a grid in which difficult regions are identified using watershed labeling.
The regions can be estimated online, e.g., based on the number of collision-free samples around a given configuration~\cite{overmarsGauss,hsuBridge}.
%The narrow passages are however located in the configuration space and it is generally not easy to estimate their location
%based only on the features of the workspace~\cite{hannaWIS}.
%.g. based on amount of collision-free samples around a given sample~\cite{overmarsGauss} or e.g. based
%~\cite{hsuBridge,overmarsGauss}.

@@

The probability of sampling of the narrow passages can be increased by dilating the free-space, e.g., by shrinking the geometry of 
the robot~\cite{hsuOnProb} or the obstacles~\cite{bayazitIRC}.
The shrinking technique has been used also in~\cite{cortes2010simulating}, where the exit pathways for a small flexible molecule are
computed using a RRT-based method.
The flexible ligand is modeled as a kinematic chain, which increases the dimension of the configuration space.
To cope with these many-DOF ligands, the RRT-ML~\cite{cortes2007mlrrt} approach is employed as the basic planner in~\cite{cortes2010simulating}.
RRT-ML expands the tree primarily using those DOF that are essential for achieving the motion of the ligand (i.e., rotation
and translation) and it employs the other DOFs (i.e., those that are responsible for conformation changes) if they hinder the growth of the tree.
In~\cite{singhLIG}, the pathways for a flexible ligand are found using Probabilistic Roadmaps.
Each configuration in the roadmap is also evaluated using electrostatic energy.
The high-energy configurations are discarded, while low-energy conformations (and their vicinity) are sampled more dense.
The approach~\cite{singhLIG} can be also used to predict ligand binding sites.

@@

Depending on the chemical interactions between the ligand and the protein surface, the ligand may enter a tunnel with various rotations and in various spatial conformations.
The traversability of the tunnel should therefore be evaluated using multiple starting configurations $\QI$.
The initial configurations are searched around the beginning of the tunnel.
A random sample $q$ is generated around the first sphere of the tunnel 
$c_1$ in the distance $\RI$ (the translation and rotation parts of $q$ are generated randomly, the scale is set to $\smin$, and the conformation index is set randomly).
If the sample $q$ is collision-free, it is added to $\QI$ as a new starting configuration.
%Similarly, a single goal configuration $\qgoal$ is found around the end of the tunnel. 
The distance $\RI$ influences how far are the initial configurations from the start of the tunnel.
By placing them too far, the planner would need more iterations to enter the tunnel, which would increase the runtime.
Contrary, too low $\RI$ may disable to find any collision-free configurations, especially for large ligands.
We suggest to set $\RI$ to the radius of the first sphere in the tunnel and increase it only if no collision-free configuration
can be found around.
For each starting configuration $\qinit \in \QI$, the trajectories are computed using a modified RRT, which is introduced in the following section.

@@

Movements of the ligands inside tunnels are blocked by many atoms forming the tunnel.
It is not always possible for the ligand to strictly follow the centerline of the tunnel and slight detours
should be allowed, which is controlled by the parameter $\dt$.
This parameter should be set considering existence of cavities or other tunnels around the tunnel being investigated.
In the case of alone tunnels (which is very common), we recommend to use double of the tunnel average width.

@@

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from the Gaussian distribution $N(c_v,\Sigma)$, where $\Sigma$ is the diagonal matrix with diagonal entries equal to the parameter $\rv$, 
and the rotational part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The ligand index $l$ and scale $s$ can be set to zero, as these are not used in the employed metric for the nearest-neighbor search.
The parameter $\rv$ influences the distribution of the random samples around the tunnel centerline. 
We propose to set this parameter to the average width of the tunnel.

%By setting $\rv$ to a small value, the planner attempts to find the trajectories inside the tunnel, while higher values
%of $\rv$ cause the exploration of paths around the tunnel.

@@

The third modification is the utilization of another metric during the expansion procedure (line~\ref{alg::expand:a} in Alg.~\ref{alg::expand}).
This metric is denoted as $\da(\cdot,\cdot)$ and it is referred to as atom-based metric in the rest of the paper.
To evaluate $\da(q_1,q_2)$, the absolute positions of the atoms of the ligand defined by $q_1$ and $q_2$ are computed.
The minimal 3D Euclidean distance from the first set of atoms (defined by $q_1$) to the second set of atoms (defined by $q_2$) is the value of 
$\da(q_1,q_2)$.
This metric is sensitive to the shape of the ligand (that depends on its conformation) as it computes the distance using the two nearest atoms of the two configurations.
It allows us to expand the tree by a configuration whose nearest atom approaches its nearest atom in $\qrand$, which supports
retraction of the ligands towards $\qrand$.

%By computing the atom-based metric using the absolute positions of atoms, this metric considers the shape of the ligands, which
%is useful when multiple conformations are utilized.
%This metric ensures that the tree is expanded by such configurations, whose atoms maximally approach
%the atoms of the configuration $\qrand$.
%We observed that this leads to a better construction of the configuration tree inside the proteins and it increases the probability of reaching the active site.

@@

The pool of available conformations $\L$ consists all conformations (upto some resolution) with energy lower than a user specified threshold, which can still contain too many conformations (thousands or even more).
We select a subset by computing how the atoms of the ligand are spread around its geometric mean.
The distance of centers of ligand atoms to the geometric center of the ligand is computed, and the deviation of these
distances then characterize the spread.
Low deviation is characteristic for the packed conformations, while larger deviations correspond to the longer conformations.
To select $n$ conformations from all $|\L|$ available conformations, they are sorted according to the deviation.
Each $k-$th conformation from the sorted list, where $k=\lfloor{|\L| / n}\rfloor$, is used.

@@

The tested methods differ in the expansion step (lines \ref{alg::expa}--\ref{alg::expb} in Alg.~\ref{alg::expand}).
In these lines, the \RB\ method employs the retraction step in order to find a new collision-free configuration $\qnew$ that approaches $\qrand$.
The retraction step is realized by sampling the vicinity of the configuration $\qnear$ in the distance $0.05$~\AA\ using 20 samples.
A collision-free sample that maximally approaches $\qrand$ is selected and added to the tree, and the retraction continues from this sample.
The retraction step is repeated 5 times.
The \RB\ method can therefore expand the tree by at most 5 new nodes per configuration, while \RA\ extends the tree only by one configuration.
The larger number of nodes in \RB\ method is not problematic, as the KD-trees were used for the nearest-neighbor search and the time to solve these queries is significantly smaller than the computing collision detection.
The parameters of the methods were selected in such a way that the number of collision detection queries in each expansion step is the same.
For each conformation and tested scale, the \RA\ method attempts to find $m=100$ collision-free configurations, and \RB\ attempts to perform $5 \times 20$ collision-detection queries.
Other parameters of the \RB\ planner were the same as for \RA.
The planning was terminated after $\Imax$ iterations or if the tree approached the active site to the distance less than 2~\AA\ (i.e., the 
geometric center of the ligand was closer than 2~\AA\ from the active site).

The performance of the methods is evaluated using the traversability rate, which is the percentage of frames of the protein dynamics that are traversable by the tested ligand.
A frame is considered traversable if at least one trajectory out of 200 ($\!|\QI|\!\times\!\!10$ trials) approaches the active site to the distance less than~2~\AA. 
This distance is measured as 3D distance between geometric center of the ligand and the position of the active site.
%~\footnote{We refer to~\url{http://mrs.felk.cvut.cz/icra2018protein} for additional material.}

@@

The performance of \RA\ and \RB\ planners is first measured by finding trajectories for only one atom of VDW radii $1.7$~\AA.
The \RC\ variant is not tested here as it behaves same as \RA\ in the case of one-atom ligand.
The computed traversability rates are shown in Tab.~\ref{tab::one}.
The maximum percentage of traversable frames (the row Maximum in Tab.~\ref{tab::one}) is computed as the percentage of tunnels with the bottleneck larger than the tested radius.
The row 'Planning' shows the traversable rate computed by the planners.
This rate decreases with the increasing radius and it is lower than the maximum traversable rate.
The rates provided by \RA\ and \RB\ are similar which is a good indicator that the methods have been implemented correctly.
Higher traversability rate can be achieved by increasing the maximum number of planning iterations $\Imax$, which would also
increase the planning time.
For the purpose of the following experiments, we used $\Imax=5,000$ as a compromise between the performance and the runtime.

@@

\begin{table}[ht]
\centering
\caption{\label{tab::selection}
    \small
    The influence of the conformation selection method on the traversability rates in the first (main) tunnel.
    The number after~`$/$' denotes the number of atoms.
    The protein has 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphSel.tex
}
\end{table}

@@

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    \small
    Traversability rate [\%] over 100 frames for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
    The protein has 4650 atoms.
}
\small
\renewcommand{\tabcolsep}{2pt}
{
\scriptsize
\renewcommand{\arraystretch}{0.7}
\input graphNew.tex
}
\end{table}

@@

The traversability rate is shown in Tab.~\ref{tab::main}. 
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the ligand DCP, trajectories were detected in 10~\% of frames for scale $\smin=0.8$ and the traversability of this ligand
to the active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
%For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (18~\%), m037t (45~\%) and m038t (30~\%).
%The traversability of the DCP ligand to active site was approved by a separate MD simulation~\cite{marques2017catalytic}.
Based on the number of success rate, also other ligands (e.g., m037t with traversability rate 45~\% at $\smin$=0.8) seem to be promising.
However, their binding was not approved yet by MD dedicated simulations.

@@

Trajectories for $\LA$ in all detected tunnels were classified as successful if they reached the last sphere of the tunnel
to distance $2$~\AA or less and they were considered unsuccessful otherwise.
The throughput computed from the trajectories shows that the tunnels are difficult not only around bottlenecks, but
also in other places.
The comparison of the classic bottleneck (i.e., measured by the radius of spherical probe) and throughput is depicted in Fig.~\ref{fig::tunnel2}a.
The trajectories for $\smin=0.4$ are depicted in Fig.~\ref{fig::tunnel2}b and for $\smin=0.5$ in Fig.~\ref{fig::tunnel2}c.
The successful trajectories for $\smin=0.4$ reveal that the end of the tunnel No. 3 can be approached by two different pathways (one in the tunnel and another one outside the tunnel).
The detail is depicted in Fig.~\ref{fig::detail}.
Despite the low bottleneck of this tunnel, the ligand may reach its end using the alternative pathway.
